---
title: "Supplementary 2 for 'Quantitative pollen-based reconstructions of climate characteristics and forest coverage for Northern Central Siberia: evaluation of different techniques'"
author: "Rodion Andreev, Elena Novenko"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---
```{r Package loading, echo=FALSE, message=FALSE, results='hide'}
library('readxl')
library('writexl')
library('tidyverse')
library('analogue')
library('rioja')
library('randomForest')
library('e1071')
library('caret')
source('R/writing_functions.R')
source('R/plotting_functions.R')
```


```{r Data loading, echo=FALSE, message=FALSE, results='hide'}
fossil = read_excel('data/fossil/fossil_NizhnyayaTunguska.xlsx') |>
  gather(variable, value, -depth) |>
  spread(depth, value) |>
  transform(variable = as.numeric(variable)) |>
  arrange(variable)
depth = fossil$variable
ages = read_excel(paste0('data/fossil/ages_NizhnyayaTunguska.xlsx'), col_names = F) |> 
  pull()

# Modern data (training set)
modern = read_excel('data/modern/modern.xlsx',
                    col_types = c('text', rep('numeric', 51))) |>
  as.data.frame() |>
  arrange(points)

# Parameters
# Woody cover
cover = read_excel('data/parameters/woody_cover.xlsx',
                   col_types = c('text', rep('numeric', 4), 'text', 'text')) |>
  as.data.frame() |>
  arrange(points)

# Climate
climate = read_excel('data/parameters/climate.xlsx',
                     col_types = c('text', rep('numeric', 4))) |> 
  arrange(points)

# Biome selection
modern = left_join(modern,
                   cover,
                   by = 'points') |>
  select(-km50:-type1) |>
  filter(!str_detect(type2, 'putorana')) |>
  select(-type2)
cover = select(cover, -type1:-type2)

# Column names switch
row.names(modern) = modern$points
row.names(fossil) = fossil$variable
colnames(fossil) = colnames(modern)

parameters_modern = semi_join(climate, modern, by = 'points') |>
  inner_join(cover)

parameters_types = colnames(parameters_modern)[-1]

fossil = select(fossil, -points)
modern = select(modern, -points)

# Excluding unnecessary taxa
fossil = fossil |> 
  select(-`Menuanthes trifoliata`, -`Cyperaceae`)
modern = modern |> 
  select(-`Menuanthes trifoliata`, -`Cyperaceae`)

recons = data.frame(depth)
```

## Scatter plots
### 1. MAT scatter plots

```{r MAT scatterplots, echo=FALSE, fig.keep='all', out.width='50%', out.height='50%'}
dat = join(modern, fossil)

# For MAT data needs to be proportional
modern_mat = dat$modern / 100 
fossil_mat = dat$fossil / 100

i = 0 # Increment
mat = data.frame(ages)
summary_mat = performance_summary(parameters_types) # for summary

for (i in 1:length(parameters_types)){
  # initial parameters
  method = 'MAT'
  parameter = pull(parameters_modern, var = parameters_types[i])
  
  # initial transfer function
  func_mat = mat(modern_mat, parameter,
                 method = "SQchord") 
  
  # bootstrap cross-validation
  cv_mat = bootstrap(func_mat, n.boot = 100)
  
  # prediction on newdata
  recon_mat = predict(func_mat, fossil_mat, k = getK(cv_mat))
  
  # pull the prediction 
  parameter1 =
    recon_mat$predictions$model$predicted[getK(cv_mat),]
  
  # bind fitted values for one parameter with others
  mat = cbind(mat, parameter1)
  colnames(mat)[length(mat)] = paste0(parameters_types[i], '.mat')
  
  # results and performance writing
  results = write_results(cv_mat) # results: residuals, predictions for plots
  parameter_name = define_name(parameters_types, i) # text name for plots
  results2 = write_screeplot(cv_mat) # pull the results for scree plots
  # performance
  model_settings = paste0('n of analogues: ', getK(cv_mat)) # settings for plots
  
  # plots
  plots(results, method, parameter_name, model_settings, is.suppl = T) |> 
    print()
}
```

Figures 1.1-1.8. Scatter plots for Modern Analogue Technique (MAT) reconstruction, observed vs. fitted. Figures 1.1-1.4 are for climatic parameters, 1.5-1.8 for woody cover. The solid black line is regression line.

### 2. WA scatter plots

```{r WA scatterplots, echo=FALSE, results='hide', fig.keep='all', out.width='50%', out.height='50%'}
condition = colSums(modern) != 0
modern_wa = modern[, condition]

i = 0 # increment
wa = data.frame(ages)
summary_wa = performance_summary(parameters_types) # for summary

for (i in 1:length(parameters_types)){
  
  # initial parameters
  method = 'WA'
  parameter = pull(parameters_modern, var = parameters_types[i])
  
  # basic transfer function
  func_wa = WA(
    modern_wa,
    parameter,
    mono = F,
    tolDW = F,
    use.N2 = F,
    tol.cut = .01,
    check.data = T,
    lean = F
  )
  
  # bootstrap cross-validation
  cv_wa = crossval(
    func_wa,
    cv.method = 'bootstrap',
    verbose = T,
    ngroups = 10,
    nboot = 100,
    h.cutoff = 0,
    h.dist = NULL
  )
  
  # prediction with newdata
  recon_wa = predict(
    func_wa,
    newdata = fossil,
    sse = F,
    nboot = 100,
    match.data = T,
    verbose = T
  )
  
  # performance & summary writing 
  parameter_name = define_name(parameters_types, i) # parameter type in text
  performance_cv = performance(cv_wa) # write performance using rioja function
  
  model_settings = names(
    performance_cv$crossval[,1][which.min(
      performance_cv$crossval[,1])]) # pull model parameters
  
  # pull fitted values: best model parameters
  parameter1 = recon_wa$fit[,model_settings]
  
  # bind fitted values for one parameter with others
  wa = cbind(wa, parameter1)
  colnames(wa)[length(wa)] = paste0(parameters_types[i], '.wa')
  
  results = write_results_rioja(cv_wa) # write results
  plots(results, method, parameter_name, model_settings, is.suppl = T) |> 
    print()
  }
```

Figures 2.1-2.8. Scatter plots for weighted averaging (WA) reconstruction, observed vs. fitted. Figures 2.1-2.4 are for climatic parameters, 2.5-2.8 for woody cover. The solid black line is regression line.

### 3. WAPLS scatter plots

```{r WAPLS scatterplots, echo=FALSE, results='hide', fig.keep='all', out.width='50%', out.height='50%'}
condition = colSums(modern) != 0
modern_wapls = modern[, condition]

i = 0 # increment
wapls = data.frame(ages)
summary_wapls = performance_summary(parameters_types) # for summary

for (i in 1:length(parameters_types)){
  
  # initial parameters
  method = 'WAPLS'
  parameter = pull(parameters_modern, var = parameters_types[i])
  
  # basic transfer function
  func_wapls = WAPLS(
    modern_wapls,
    parameter,
    npls = 3,
    iswapls = T,
    standx = F,
    lean = F,
    check.data = T
  )
  
  # bootstrap cross-validation
  cv_wapls = crossval(
    func_wapls,
    cv.method = 'bootstrap',
    verbose = T,
    ngroups = 10,
    nboot = 100
  )
  
  # prediction with newdata
  recon_wapls = predict(
    func_wapls,
    newdata = fossil,
    sse = F,
    nboot = 100,
    match.data = T,
    verbose = T
  )
  
  # performance & summary writing 
  parameter_name = define_name(parameters_types, i) # parameter type in text
  performance_cv = performance(cv_wapls) # write performance 
  
  model_settings = names(
    performance_cv$crossval[,1][which.min(
      performance_cv$crossval[,1])]) # pull model parameters
  
  # pull fitted values: best parameters
  parameter1 = recon_wapls$fit[,model_settings]
  
  # bind fitted values for one parameter with others
  wapls = cbind(wapls, parameter1)
  colnames(wapls)[length(wapls)] = paste0(parameters_types[i], '.wapls')
  
  results = write_results_rioja(cv_wapls) # write results
  plots(results, method, parameter_name, model_settings, is.suppl = T) |> 
    print()
}
```

Figures 3.1-3.8. Scatter plots for partial least squares (WAPLS) reconstruction, observed vs. fitted. Figures 3.1-3.4 are for climatic parameters, 3.5-3.8 for woody cover. The solid black line is regression line.

### 4. RF scatter plots

```{r RF scatterplots, echo=FALSE, results='hide', fig.keep='all', out.width='50%', out.height='50%', warning=FALSE}
condition = colSums(modern) != 0
modern_rf = modern[, condition]

# Column (taxa) names renaming for randomForest()
names = c(1:length(modern))
i = 0
for (i in 1:length(names)) {
  names[i] = paste0('T', names[i])
}
modern_rf = modern
colnames(modern_rf) = names
fossil_rf = fossil
colnames(fossil_rf) = names

# Variables meaning
meaning = data.frame(names, colnames(modern))

# 3. Transfer function ----
# Transfer function evaluation using k-fold cross-validation. caret train()
# method is being used. RF iterates three times in order to evaluate maxnodes,
# ntree, and mtry RF parameters. Also performance results are being written
# into text file
i = 0
rf = data.frame(ages)
method = 'RF'
summary_rf = performance_summary(parameters_types) # for summary

# CV parameters
params_cv = trainControl(method = 'cv', number = 5, search = 'grid')

# Cycle repeats for each reconstruction parameter
for (i in 1:length(parameters_types)){
  
  # Data preparation
  parameter = pull(parameters_modern, var = parameters_types[i])
  dataset = cbind(parameter, modern_rf)
  
  # Grid for the first CV iteration
  grid = expand.grid(.mtry = c(1:20))
  
  # First RF iterarion: the best mtry
  cv_rf_mtry = train(
    parameter ~ .,
    data = dataset,
    method = 'rf',
    metric = 'Rsquared',
    tuneGrid = grid,
    trControl = params_cv,
    importance = T,
    nodesize = 14,
    ntree = 300
  )
  best_mtry = as.numeric(cv_rf_mtry$bestTune$mtry)
  
  # Second iteration: best maxnodes
  # New search grid using new mtry value
  grid = expand.grid(.mtry = best_mtry)
  maxnodes_storage = list()
  for (maxnodes in c(10:30)) {
    set.seed(1234)
    cv_rf_maxnode = train(
      parameter ~ .,
      data = dataset,
      method = 'rf',
      metric = 'Rsquared',
      tuneGrid = grid,
      trControl = params_cv,
      importance = T,
      nodesize = 14,
      maxnodes = maxnodes,
      ntree = 300
    )
    current_iteration = toString(maxnodes)
    maxnodes_storage[[current_iteration]] = cv_rf_maxnode
  }
  
  # Writing the results 
  results_maxnodes = resamples(maxnodes_storage)
  maxnodes_colname = colnames(
    results_maxnodes$values)[apply(
      results_maxnodes$values, 1, which.max)][1]
  best_maxnodes = as.numeric(substr(maxnodes_colname, 1, 2))
  
  # Third iterarion: the best ntree
  ntree_storage = list()
  for (ntree in c(250, 300, 350, 400, 450, 500, 550, 600, 800, 1000, 2000)){
    set.seed(1234)
    cv_rf_ntree = train(
      parameter ~ .,
      data = dataset,
      method = 'rf',
      metric = 'Rsquared',
      tuneGrid = grid,
      trControl = params_cv,
      importance = T,
      nodesize = 14,
      maxnodes = best_maxnodes,
      ntree = ntree
    )
    ntree_value = toString(ntree)
    ntree_storage[[ntree_value]] = cv_rf_ntree
  }
  
  # Writing the results 
  results_ntree = resamples(ntree_storage)
  values = results_ntree$values
  ntree_colname = colnames(values)[apply(values, 1, which.max)][1]
  best_ntree = as.numeric(substr(ntree_colname, 1, 3))
  
  # Returning R2 values
  r2_flt = str_detect(colnames(values), 'Rsquared')
  r2_rf = round(apply(values[r2_flt], 1, max)[1], 2)
  
  # Returning RMSEP values
  rmsep_flt = str_detect(colnames(values), 'RMSE')
  rmsep_rf = round(apply(values[rmsep_flt], 1, min)[1], 2)
  
  # Final reconstruction
  func_rf = train(
    parameter ~ .,
    data = dataset,
    method = "rf",
    metric = "Rsquared",
    tuneGrid = grid,
    trControl = params_cv,
    importance = T,
    nodesize = 14,
    ntree = best_ntree,
    maxnodes = best_maxnodes
  )
  
  # Reconstruction results 
  recon_rf = predict(func_rf, fossil_rf)
  rf = cbind(rf, recon_rf)
  colnames(rf)[length(rf)] = paste0(parameters_types[i], '.rf')
  parameter_name = define_name(parameters_types, i)
  
  # Scatter plots
  results = write_results_rf(func_rf)
  plots(
    results,
    method,
    parameter_name,
    paste0(
      'ntree = ',
      best_ntree,
      ', mtry = ',
      best_mtry,
      ', maxnodes = ',
      best_maxnodes
    ),
    is.suppl = T
  ) |> 
    print()
}
```

Figures 4.1-4.8. Scatter plots for Random Forest (RF) reconstruction, observed vs. fitted. Figures 4.1-4.4 are for climatic parameters, 4.5-4.8 for woody cover. The solid black line is regression line.
