---
title: "Supplementary 1 for ..."
author: "Rodion A. Andreev"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
```{r Package loading, echo=FALSE, message=FALSE, results='hide'}
library('readxl')
library('writexl')
library('tidyverse')
library('analogue')
library('rioja')
library('randomForest')
library('e1071')
library('caret')
```


```{r Data loading, echo=FALSE, message=FALSE, results='hide'}
fossil = read_excel('fossil/fossil_Gornoe.xlsx') |> 
  gather(variable, value, -depth) |>  
  spread(depth, value) |>
  transform(variable = as.numeric(variable)) |> 
  arrange(variable)

# Modern data (training set)
modern = read_excel('modern/modern_new.xlsx', 
                    col_types = c('text', rep('numeric', 51))) |> 
  as.data.frame() |> 
  arrange(points)

# Parameters
# Woody cover
cover = read_excel('parameters/woody_cover.xlsx',
                   col_types = c('text', rep('numeric', 4), 'text', 'text')) |> 
  as.data.frame() |> 
  arrange(points)

# Climate
climate = read_excel('parameters/climate.xlsx',
                     col_types = c('text', rep('numeric', 4))) |> 
  arrange(points)

# Biome selection
modern = left_join(modern, 
                   cover,
                   by = 'points') |> 
  select(-km50:-type1) |> 
  filter(!str_detect(type2, 'putorana')) |> 
  select(-type2)
cover = select(cover, -type1:-type2)

row.names(modern) = modern$points
row.names(fossil) = fossil$variable
colnames(fossil) = colnames(modern)

parameters_modern = semi_join(climate, modern, by = 'points') |>
  inner_join(cover)

parameters_types = colnames(parameters_modern)[-1]

fossil = select(fossil, -points)
modern = select(modern, -points)

depth = rownames(fossil)
ages = read_excel('fossil/ages_Gornoe.xlsx', col_names = F) |> 
  pull()
recons = data.frame(depth)

write_results = function(t_function){
  results = tibble(
    fitted_recon = fitted(t_function)$estimated,
    residuals_recon = resid(t_function, k = getK(t_function))$bootstrap[,getK(t_function)],
    observed_orig = t_function$observed
  )
}

write_results_rioja = function(cv){
  performance = performance(cv)
  type = names(performance[["crossval"]][,1][which.min(performance[["crossval"]][,1])])
  fitted_all = as.data.frame(fitted(cv))
  residuals_all = as.data.frame(residuals(cv, cv = T))
  results = tibble(
    fitted_recon = fitted_all[, type],
    residuals_recon = residuals_all[, type],
    observed_orig = cv[['x']]
  )
  return(results)
}

write_results_rf = function(t_function){
  results = tibble(
    fitted_recon = t_function$finalModel$predicted,
    residuals_recon = t_function$finalModel$y - t_function$finalModel$predicted,
    observed_orig = t_function$finalModel$y
  )
  return(results)
}

plot2 = function(dataset, method_type, param_name, settings, number){
  regr_line1 = tibble(
    fitted = fitted(lm(fitted_recon ~ observed_orig,
                       data = dataset)),
    observed = dataset$observed_orig
  )
  
  plot = ggplot(data = dataset) +
    geom_line(data = regr_line1,
              aes(x = observed,
                  y = fitted),
              color = 'black',
              linewidth = 0.2) +
    geom_point(mapping = aes(x = observed_orig,
                             y = fitted_recon),
               color = 'salmon',
               alpha = 0.5,
               size = 3,
               shape = 1) +
    ggtitle(paste0('Fig. ', toString(number), ', ', method_type),
            subtitle = paste0(param_name, ', ', settings)) +
    xlab('Obsevrved') +
    ylab('Fitted') +
    theme_classic()
  
  return(plot)
}

define_name = function(x, i){
  if (x[i] == 'T_jan'){
    param_name = 'January mean T, °C'
  } else if (x[i] == 'T_jul'){
    param_name = 'July mean T, °C'
  } else if (x[i] == 'T_ann'){
    param_name = 'Annual mean T, °C'
  } else if (x[i] == 'P_ann'){
    param_name = 'Annual precipitation, mm'
  } else if (x[i] == 'km50'){
    param_name = 'Woody cover, 50 km radius'
  } else if (x[i] == 'km20'){
    param_name = 'Woody cover, 20 km radius'
  } else if (x[i] == 'km10'){
    param_name = 'Woody cover, 10 km radius'
  } else if (x[i] == 'km5'){
    param_name = 'Woody cover, 5 km radius'
  }
  return(param_name)
}
```


### MAT scatterplots

```{r MAT scatterplots, echo=FALSE, results='hide', fig.keep='all', out.width='50%', out.height='50%'}
dat = join(modern, fossil, verbose = T)
modern_mat = dat$modern / 100 
fossil_mat = dat$fossil / 100
fossil_mat = fossil_mat |> 
  select(-`Menuanthes trifoliata`)
modern_mat = modern_mat |> 
  select(-`Menuanthes trifoliata`)
set.seed(1234)

i = 0
mat = data.frame(ages)

for (i in 1:length(parameters_types)){
  method = 'MAT'
  
  parameter = pull(parameters_modern, var = parameters_types[i])
  
  func_mat = mat(modern_mat, parameter,
                 method = "SQchord")
  cv_mat = bootstrap(func_mat, n.boot = 100)
  recon_mat = predict(func_mat, fossil_mat, k = getK(cv_mat))
  parameter1 = recon_mat[["predictions"]][["model"]][["predicted"]][getK(cv_mat),]
  mat = cbind(mat, parameter1)
  colnames(mat)[length(mat)] = paste0(parameters_types[i], '.mat')
  
  results = write_results(cv_mat)
  parameter_name = define_name(parameters_types, i)
  model_settings = paste0('n of analogues: ', getK(cv_mat))
  
  print(plot2(results, method, parameter_name, model_settings, i))
}
```

Figures 1.1-1.8. Scatterplots for Modern Analogue Technique (MAT) reconstruction, observed vs. fitted. Figures 1.1-1.4 are for climatic parameters, 1.5-1.8 for woody cover. The solid black line is regression line.

### WA scatterplots

```{r WA scatterplots, echo=FALSE, results='hide', fig.keep='all', out.width='50%', out.height='50%'}
condition = colSums(modern) != 0
modern_wa = modern[, condition]

i = 0
wa = data.frame(ages)

for (i in 1:length(parameters_types)){
  method = 'WA'

  parameter = pull(parameters_modern, var = parameters_types[i])

  func_wa = WA(modern_wa, parameter,
               mono = F, tolDW = F, use.N2 = F, tol.cut=.01,
               check.data = T, lean = F)

  cv_wa = crossval(func_wa, cv.method = 'bootstrap', verbose = T, ngroups = 10,
                   nboot = 100, h.cutoff = 0, h.dist = NULL)

  recon_wa = predict(func_wa, newdata = fossil, sse = F, nboot = 100,
                     match.data = T, verbose = T)

  parameter1 = recon_wa$fit[,1] |>
    as.data.frame()
  wa = cbind(wa, parameter1)
  colnames(wa)[length(wa)] = paste0(parameters_types[i], '.wa')

  parameter_name = define_name(parameters_types, i)
  performance_cv = performance(cv_wa)
  model_settings = names(performance_cv[["crossval"]][,1][which.min(performance_cv[["crossval"]][,1])])
  results = write_results_rioja(cv_wa)
  print(plot2(results, method, parameter_name, model_settings, i))
 }
```

Figures 2.1-2.8. Scatterplots for weighted averaging (WA) reconstruction, observed vs. fitted. Figures 2.1-2.4 are for climatic parameters, 2.5-2.8 for woody cover. The solid black line is regression line.

### WAPLS scatterplots

```{r WAPLS scatterplots, echo=FALSE, results='hide', fig.keep='all', out.width='50%', out.height='50%'}
condition = colSums(modern) != 0
modern_wapls = modern[, condition]

i = 0
wapls = data.frame(ages)
for (i in 1:length(parameters_types)){
  method = 'WAPLS'

  parameter = pull(parameters_modern, var = parameters_types[i])

  func_wapls = WAPLS(modern_wapls, parameter, npls = 3, iswapls = T, standx = F,
                     lean = F, check.data = T)

  cv_wapls = crossval(func_wapls, cv.method = 'bootstrap', verbose = T, ngroups=10,
                      nboot = 100)

  recon_wapls = predict(func_wapls, newdata = fossil, sse = F, nboot=100,
                        match.data = T, verbose = T)

  parameter1 = recon_wapls[["fit"]] |>
    as.data.frame() |>
    select(Comp03)
  wapls = cbind(wapls, parameter1)
  colnames(wapls)[length(wapls)] = paste0(parameters_types[i], '.wapls')

  parameter_name = define_name(parameters_types, i)
  performance_cv = performance(cv_wapls)
  model_settings = names(performance_cv[["crossval"]][,1][which.min(performance_cv[["crossval"]][,1])])
  results = write_results_rioja(cv_wapls)
  print(plot2(results, method, parameter_name, model_settings, i))
}
```

Figures 3.1-3.8. Scatterplots for partial least squares (WAPLS) reconstruction, observed vs. fitted. Figures 3.1-3.4 are for climatic parameters, 3.5-3.8 for woody cover. The solid black line is regression line.

### RF scatterplots

```{r RF scatterplots, echo=FALSE, results='hide', fig.keep='all', out.width='50%', out.height='50%', warning=FALSE}
condition = colSums(modern) != 0
modern_rf = modern[, condition]

names = c(1:length(modern))
i = 0
for (i in 1:length(names)) {
  names[i] = paste0('T', names[i])
}
modern_rf = modern
colnames(modern_rf) = names
fossil_rf = fossil
colnames(fossil_rf) = names

i = 0
rf = data.frame(ages)
params_cv = trainControl(method = 'cv', number = 5, search = 'grid')
method = 'RF'

for (i in 1:length(parameters_types)){
  parameter = pull(parameters_modern, var = parameters_types[i])
  dataset = cbind(parameter, modern_rf)
  grid = expand.grid(.mtry = c(1:20))
  
  cv_rf_mtry = train(parameter~., data = dataset,
                     method = 'rf',
                     metric = 'Rsquared',
                     tuneGrid = grid,
                     trControl = params_cv, 
                     importance = T,
                     nodesize = 14,
                     ntree = 300)
  best_mtry = as.numeric(cv_rf_mtry$bestTune$mtry)

  grid = expand.grid(.mtry = best_mtry)
  maxnodes_storage = list()
  for (maxnodes in c(10:30)){
    set.seed(1234)
    cv_rf_maxnode = train(parameter~., data = dataset,
                          method = 'rf',
                          metric = 'Rsquared',
                          tuneGrid = grid,
                          trControl = params_cv, 
                          importance = T,
                          nodesize = 14,
                          maxnodes = maxnodes,
                          ntree = 300)
    current_iteration = toString(maxnodes)
    maxnodes_storage[[current_iteration]] = cv_rf_maxnode
  }
  results_maxnodes = resamples(maxnodes_storage)
  maxnodes_colname = colnames(results_maxnodes$values)[apply(results_maxnodes$values, 1, which.max)][1]
  best_maxnodes = as.numeric(substr(maxnodes_colname, 1, 2))
  
  ntree_storage = list()
  for (ntree in c(250, 300, 350, 400, 450, 500, 550, 600, 800, 1000, 2000)){
    set.seed(1234)
    cv_rf_ntree = train(parameter~., data = dataset,
                        method = 'rf',
                        metric = 'Rsquared',
                        tuneGrid = grid,
                        trControl = params_cv, 
                        importance = T,
                        nodesize = 14,
                        maxnodes = best_maxnodes,
                        ntree = ntree)
    ntree_value = toString(ntree)
    ntree_storage[[ntree_value]] = cv_rf_ntree
  }
  results_ntree = resamples(ntree_storage)
  summary(results_ntree)
  values = results_ntree$values
  ntree_colname = colnames(values)[apply(values, 1, which.max)][1]
  best_ntree = as.numeric(substr(ntree_colname, 1, 3))
  
  r2_rf = apply(values, 2, max)[colnames(values)[apply(values, 1, which.max)][1]] |> 
    as.numeric() |> 
    round(2)
  
  rmsep_flt = str_detect(colnames(values), 'RMSE')
  rmsep_rf = round(apply(values[rmsep_flt], 1, min)[1], 2)
  
  func_rf = train(parameter~., data = dataset,
                   method = "rf",
                   metric = "Rsquared",
                   tuneGrid = grid,
                   trControl = params_cv,
                   importance = T,
                   nodesize = 14,
                   ntree = best_ntree,
                   maxnodes = best_maxnodes)
  
  recon_rf = predict(func_rf, fossil_rf)
  rf = cbind(rf, recon_rf)
  colnames(rf)[length(rf)] = paste0(parameters_types[i], '.rf')
  parameter_name = define_name(parameters_types, i)
  results = write_results_rf(func_rf)
  
  print(plot2(results, method, parameter_name, paste0('ntree = ', best_ntree, ',
                                                mtry = ', best_mtry,
                                                ', maxnodes = ', best_maxnodes),
        i))
}
```

Figures 4.1-4.8. Scatterplots for Random Forest (RF) reconstruction, observed vs. fitted. Figures 4.1-4.4 are for climatic parameters, 4.5-4.8 for woody cover. The solid black line is regression line.
